Two subtle problems with over-representation analysis

### 2024-07-18 manuscript/manuscript.Rmd
downsampling added to ms

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results.
ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

>>>>

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
+selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

## Does FCS have better sensitivity than ORA with real RNA-seq data?

To understand whether FCS has better sensitivity than ORA in a real setting, we conducted a
systematic downsampling of a gene expression dataset cosisting of normal and tumor samples
of 37 lung cancer patients.
In the full set of patient samples, fora (ORA) yielded 375 significantly differentially
regulated Reactomes, while fgsea (FCS) gave 470 (FDR<0.05).
Fgsea yielded consistently more significant pathways than fora in downsampled data
(**Figure 4A**).
Fgsea also identified a larger proportion of pathways in downsampled data compared to fora.
For example at a sample size of 5, fgsea identified 74.5% of pathways from the full set as
compared to 41.5% for fora, confirming better recall for FCS over ORA.
It was noted that fgsea identified more pathways in downsampled data that were
inconsistent with the full dataset as compared to fora (**Figure 4B**).
However, inconsistent pathways were a smaller proportion of all findings for fgsea as
compared to fora (**Figure 4C**), indicating an overall lower false positive rate for FCS
compared to ORA methods.

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.
(A) The number of significant pathways identified in downsampled data that are consistent
with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
(B) The number of significant pathways identified in downsampled data that are not
consistent the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.
All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed with
downsampled real cancer RNA-seq data means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results.
ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
